---
title: "hw5gty"
author: "Tianyue Guo"
date: "2024-11-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Refer to some ways of language grammar in ChatGpt.


## Q1

```{r}
library(data.table) # for fread

# Load digits.csv
digits_path <- "digits.csv"

# Function to load and preprocess digits data
load_digits <- function(subset=NULL, normalize=TRUE) {
  df <- fread(digits_path)
  df <- as.matrix(df)

  # Only keep the numbers we want
  if (length(subset) > 0) {
    c <- dim(df)[2]
    l_col <- df[, c]
    index = NULL

    for (i in 1:length(subset)) {
      number = subset[i]
      index = c(index, which(l_col == number))
    }
    index <- sort(index)
    df <- df[index, ]
  }

  # Convert to arrays
  digits <- df[, -1]
  labels <- df[, c]

  # Normalize digit values to 0 and 1
  if (normalize == TRUE) {
    digits <- digits - min(digits)
    digits <- digits / max(digits)
  }

  # Change the labels to 0 and 1
  for (i in 1:length(subset)) {
    labels[labels == subset[i]] <- i - 1
  }

  return(list(digits = digits, labels = labels))
}

# Function to split data into training and testing sets
split_samples <- function(digits, labels) {
  num_samples <- dim(digits)[1]
  num_training <- round(num_samples * 0.7)
  indices <- sample(1:num_samples, size = num_samples)
  training_idx <- indices[1:num_training]
  testing_idx <- indices[-(1:num_training)]

  return(list(
    training_digits = digits[training_idx, ],
    training_labels = labels[training_idx],
    testing_digits = digits[testing_idx, ],
    testing_labels = labels[testing_idx]
  ))
}

# Linear SVM implementation
train_svm <- function(training_digits, training_labels, C) {
  n <- nrow(training_digits)
  p <- ncol(training_digits)

  # Initialize weights and bias
  weights <- rep(0, p)
  bias <- 0

  # Learning rate
  eta <- 0.01

  # Number of iterations
  max_iter <- 1000

  for (iter in 1:max_iter) {
    for (i in 1:n) {
      xi <- training_digits[i, ]
      yi <- ifelse(training_labels[i] == 1, 1, -1)

      # Check the condition for misclassification
      if (yi * (sum(weights * xi) + bias) < 1) {
        weights <- weights + eta * (yi * xi - 2 * C * weights)
        bias <- bias + eta * yi
      } else {
        weights <- weights - eta * 2 * C * weights
      }
    }
  }

  return(list(weights = weights, bias = bias))
}

# Prediction function for SVM
predict_svm <- function(digits, model) {
  weights <- model$weights
  bias <- model$bias

  predictions <- as.numeric(digits %*% weights + bias > 0)
  return(predictions)
}

# Accuracy function
calculate_accuracy <- function(predictions, labels) {
  return(mean(predictions == labels))
}

# Main program
set.seed(42)
result <- load_digits(subset = c(1, 7), normalize = TRUE)
digits <- result$digits
labels <- result$labels

split_result <- split_samples(digits, labels)
training_digits <- split_result$training_digits
training_labels <- split_result$training_labels
testing_digits <- split_result$testing_digits
testing_labels <- split_result$testing_labels

# Train linear SVM with C = 1
C <- 1
svm_model <- train_svm(training_digits, training_labels, C)

# Predict on training and testing data
train_predictions <- predict_svm(training_digits, svm_model)
test_predictions <- predict_svm(testing_digits, svm_model)

# Calculate accuracy
train_accuracy <- calculate_accuracy(train_predictions, training_labels)
test_accuracy <- calculate_accuracy(test_predictions, testing_labels)

# Print results
cat("Training Accuracy:", train_accuracy, "\n")
cat("Testing Accuracy:", test_accuracy, "\n")

```


I find that the SVM model achieve high accuracy on both the training set and the test set, which are very close to each other. I think this means that the model has good learning ability and the data sample data may be linearly separable.


## Q2

(a)

To dive into linearly non-separable data set, I introduced kernel to the SVM model and first choose different C to control slackness.

```{r}


# Kernel SVM implementation
train_kernel_svm <- function(training_digits, training_labels, kernel_func, C) {
  n <- nrow(training_digits)

  # Initialize alphas and bias
  alphas <- rep(0, n)
  bias <- 0

  # Precompute the kernel matrix
  kernel_matrix <- matrix(0, n, n)
  for (i in 1:n) {
    for (j in 1:n) {
      kernel_matrix[i, j] <- kernel_func(training_digits[i, ], training_digits[j, ])
    }
  }

  # Learning rate
  eta <- 0.01

  # Number of iterations
  max_iter <- 1000

  for (iter in 1:max_iter) {
    for (i in 1:n) {
      yi <- ifelse(training_labels[i] == 1, 1, -1)
      prediction <- sum(alphas * training_labels * kernel_matrix[, i]) + bias

      # Update alphas and bias
      if (yi * prediction < 1) {
        alphas[i] <- alphas[i] + eta * (1 - yi * prediction)
        bias <- bias + eta * yi
      }
    }
  }

  return(list(alphas = alphas, bias = bias, kernel_func = kernel_func, training_digits = training_digits, training_labels = training_labels))
}

# Prediction function for Kernel SVM
predict_kernel_svm <- function(digits, model) {
  alphas <- model$alphas
  bias <- model$bias
  kernel_func <- model$kernel_func
  training_digits <- model$training_digits
  training_labels <- model$training_labels

  n <- nrow(digits)
  predictions <- rep(0, n)

  for (i in 1:n) {
    prediction <- sum(alphas * training_labels * sapply(1:length(alphas), function(j) kernel_func(digits[i, ], training_digits[j, ]))) + bias
    predictions[i] <- ifelse(prediction > 0, 1, 0)
  }

  return(predictions)
}

# Main program for Task 2
set.seed(42)
result <- load_digits(subset = c(1, 7), normalize = TRUE)
digits <- result$digits
labels <- result$labels

split_result <- split_samples(digits, labels)
training_digits <- split_result$training_digits
training_labels <- split_result$training_labels
testing_digits <- split_result$testing_digits
testing_labels <- split_result$testing_labels

# Experiment with different C values
C_values <- c(0.1, 1, 10)
for (C in C_values) {
  svm_model <- train_svm(training_digits, training_labels, C)
  train_predictions <- predict_svm(training_digits, svm_model)
  test_predictions <- predict_svm(testing_digits, svm_model)

  train_accuracy <- calculate_accuracy(train_predictions, training_labels)
  test_accuracy <- calculate_accuracy(test_predictions, testing_labels)

  cat("C =", C, "Training Accuracy:", train_accuracy, "Testing Accuracy:", test_accuracy, "\n")
}
```
From the results, I find that higher C values may increase margin but introduce more errors. Lower C values can reduce errors, taking smaller margin as a cost.

(b)

Then I try kernel SVM with polynomial kernel.

```{r}
# Define a custom kernel (e.g., polynomial kernel)
polynomial_kernel <- function(x, y, degree = 2) {
  return((sum(x * y) + 1)^degree)
}

# Train and evaluate Kernel SVM
kernel_model <- train_kernel_svm(training_digits, training_labels, polynomial_kernel, C = 1)
kernel_train_predictions <- predict_kernel_svm(training_digits, kernel_model)
kernel_test_predictions <- predict_kernel_svm(testing_digits, kernel_model)

kernel_train_accuracy <- calculate_accuracy(kernel_train_predictions, training_labels)
kernel_test_accuracy <- calculate_accuracy(kernel_test_predictions, testing_labels)

cat("Kernel SVM Training Accuracy:", kernel_train_accuracy, "\n")
cat("Kernel SVM Testing Accuracy:", kernel_test_accuracy, "\n")

```

The Polynomial kernel SVM doesn't have a very high accuracy since the training and testing one are not very close to each other. Therefore, the origin SVM is a better choice.  