---
title: "hw2"
author: "Tianyue Guo"
date: "2024-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Refer to some ways of language grammar in ChatGpt.

## Step 1: Reading and Cleaning the Data



```{r}

# Load necessary libraries
library(tidyverse)


# Read the data without interpreting any column as row names
data <- read.table("LAhousingpricesaug2013.txt", header = TRUE, na.strings = "n/a")

# Select relevant columns and clean data
data_cleaned <- data %>%
  select(SalesofSingleFamilyHomes, 
         PriceMedianSFR.1000., 
         PriceMedianCondos.1000., 
         MedianHomePrice.Sq.Ft) %>%
  drop_na()

# Extract Y and X variables
Y <- data_cleaned$SalesofSingleFamilyHomes
X1 <- data_cleaned$PriceMedianSFR.1000.
X2 <- data_cleaned$PriceMedianCondos.1000.
X3 <- data_cleaned$MedianHomePrice.Sq.Ft

# Create the design matrix X (with intercept)
X <- cbind(1, X1, X2, X3)

cat("Cleaned data has", nrow(data_cleaned), "observations.\n")
print(X)
```

## Step 2: 1）Linear Regression with Gauss-Jordan Elimination

```{r}
gauss_jordan <- function(X, Y) {
  # Create the augmented matrix [X'X | X'Y]
  XtX <- t(X) %*% X
  XtY <- t(X) %*% Y
  A <- cbind(XtX, XtY)  # Augmented matrix
  
  n <- nrow(A)  # Number of rows
  
  # Perform Gauss-Jordan elimination with pivoting
  for (i in seq_len(n)) {
    # Pivot step: Find the row with the largest value in the current column
    pivot_row <- which.max(abs(A[i:n, i])) + (i - 1)
    A[c(i, pivot_row), ] <- A[c(pivot_row, i), ]  # Swap rows

    # Normalize the pivot row
    A[i, ] <- A[i, ] / A[i, i]

    # Eliminate other rows
    for (j in seq_len(n)) {
      if (i != j) {
        A[j, ] <- A[j, ] - A[j, i] * A[i, ]
      }
    }
  }

  # Extract the solution (β coefficients)
  beta <- A[, ncol(A)]
  return(beta)
}

# Compute β coefficients using Gauss-Jordan elimination
beta_gj <- gauss_jordan(X, Y)
cat("Estimated β using Gauss-Jordan Elimination:", beta_gj, "\n")
```

## 2) Using sweep operator

```{r}
sweep_operator <- function(X, Y) {
  # Create the augmented matrix [X'X | X'Y]
  XtX <- t(X) %*% X
  XtY <- t(X) %*% Y
  A <- cbind(XtX, XtY)  # Augmented matrix

  n <- ncol(X)  # Number of predictors including intercept

  # Apply the sweep operator
  for (k in seq_len(n)) {
    A[k, ] <- A[k, ] / A[k, k]
    for (i in seq_len(n)) {
      if (i != k) {
        A[i, ] <- A[i, ] - A[i, k] * A[k, ]
      }
    }
  }

  # Extract the solution (β coefficients)
  beta <- A[, ncol(A)]
  return(beta)
}

# Compute β coefficients using the Sweep Operator
beta_sweep <- sweep_operator(X, Y)
cat("Estimated β using Sweep Operator:", beta_sweep, "\n")

```

## Step 3 & 4: Leave-One-Out Regression

```{r}
# Initialize matrix to store β^(−i) for each iteration
beta_minus_i <- matrix(NA, nrow = nrow(X), ncol = ncol(X))

# Perform leave-one-out regression for each row
for (i in seq_len(nrow(X))) {
  # Remove the i-th observation
  X_minus_i <- X[-i, ]
  Y_minus_i <- Y[-i]

  # Compute β^(−i) using Gauss-Jordan Elimination
  beta_i <- tryCatch({
    gauss_jordan(X_minus_i, Y_minus_i)
  }, error = function(e) {
    cat("Error for observation", i, ":", e$message, "\n")
    rep(NA, ncol(X))  # Return NA if an error occurs
  })

  # Store the result in the matrix
  beta_minus_i[i, ] <- beta_i
}

# Print the β^(−i) matrix
cat("Leave-One-Out β coefficients:\n")
print(beta_minus_i)

```

## Step 5: Plot the Influence of Each Observation

```{r}
# Load the ggplot2 library
library(ggplot2)

# Create a data frame for plotting
influence_data <- data.frame(
  Observation = 1:nrow(beta_minus_i),
  Beta_1 = beta_minus_i[, 2]  # Extract the β1^(−i) values
)

# Generate the plot using ggplot2
ggplot(influence_data, aes(x = Observation, y = Beta_1)) +
  geom_line(color = "blue") +
  geom_point(color = "red", size = 2) +
  labs(
    title = expression("Influence of Each Observation on " ~ hat(beta)[1]),
    x = "Observation Index (i)",
    y = expression(hat(beta)[1]^{(-i)})
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

The plot shows that $\beta_{1}^{(-i)}$ is relatively stable across most observations, except for several observations, where we observe some extremely high or low values, close to 0.0067 and 0.01. This indicates that these observations have a moderate influence on the model.